---
title: "Food Prices and Inflation"
subtitle: "Stat 184"
author:
  - "Hanna Oh"
  - "Aria Hooybergs"
  - "Aryan Deshmukh"
date: "May 6, 2026"
date-modified: now

format:
  pdf:
    documentclass: article
    toc: true
    number-sections: true
    colorlinks: true
    geometry:
      - top=0.8in
      - bottom=0.8in
      - left=0.7in
      - right=0.7in

execute:
  echo: false
  warning: false
  message: false

editor: visual
bibliography: references.bib
csl: apa.csl
---

# Introduction

Our project explores how the average prices of selected food items have changed over time in U.S. cities [@bls_prices].

We chose this topic because we are interested in economic trends such as inflation and the cost of living. By looking at everyday food items, we can better understand how prices change over time and how those changes may affect daily life.

Our main goal is to identify trends and patterns in food price changes and to examine how the prices of common food items have shifted over the years within an inflation context.

Initially, we planned to compare broader CPI categories, such as food at home and food away from home. However, after exploring the available data, we adjusted our focus to individual food items because the dataset we used provided detailed average price information rather than category-level CPI indices.

# Data

We used publicly available data from the U.S. Bureau of Labor Statistics (BLS) [@bls_prices]. Our main dataset was the Average Price dataset, which provides monthly price data for individual food items across U.S. cities. We also used an additional dataset focusing specifically on Grade A egg prices and their monthly percent changes [@bls_eggs].

For the Average Price dataset, the main variables used in our analysis include the time period, food item, and average price value. For the egg dataset, the main variables include date and monthly percent change. In our analysis, each case represents a time-point observation for a specific food item.

# Data Provenance

The data were collected and published by the U.S. Bureau of Labor Statistics, a federal agency responsible for measuring labor market activity, prices, and other economic conditions in the United States.

The Average Price dataset reports monthly price information for selected consumer goods, including food items, across U.S. cities. These data are intended to help track changes in consumer prices over time and support public understanding of inflation and cost-of-living trends.

The data are publicly available through the BLS website and are provided in downloadable tabular formats. Because the datasets are maintained by a government statistical agency, they include documentation and standardized variable structures, which support reproducible analysis.

# FAIR and CARE Principles

## FAIR Principles

The datasets align well with the FAIR principles.

-   **Findable:** The data are publicly available through the U.S. Bureau of Labor Statistics website and are accompanied by documentation.
-   **Accessible:** The files can be downloaded directly and used in statistical software such as R.
-   **Interoperable:** The data are stored in tabular formats that can be imported, cleaned, and combined with other datasets.
-   **Reusable:** Because the data are well-documented and publicly distributed, they can be reused for future analysis.

## CARE Principles

The datasets align more clearly with FAIR principles than CARE principles. CARE principles are especially important for data involving Indigenous communities or community-governed data. In this case, the BLS data are public government datasets and do not specifically involve those kinds of community-based data concerns.

However, the data still support some CARE-related values in a broader sense. They contribute to public understanding of economic conditions and can help people better understand trends related to prices, inflation, and daily living costs.

# Data Cleaning and Wrangling

Before analyzing the data, we needed to clean the raw datasets because they were not immediately ready for analysis.

For the Average Price dataset, we imported multiple files and merged them into one complete dataset. We then converted the price values into numeric format, created a time variable using the year and month, selected a set of representative food items, and removed missing values.

For the egg price dataset, we cleaned the data by formatting the date variable, converting percentage change values into numeric format, and removing missing or invalid entries.

We used R and tidyverse tools, including dplyr, tidyr, ggplot2, and readr, to complete  steps and prepare clean datasets for analysis.

# Exploratory Data Analysis
  
```{r, include=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)

food_clean <- read_csv("clean_food_prices.csv")

food_clean <- food_clean |>
  mutate(date = as.Date(date))
```

The line plot below shows changes in average prices of selected food items over time.

```{r}
#| echo: false
#| eval: true
ggplot(food_clean, aes(x = date, y = value, color = item_clean)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Average Prices of Selected Food Items Over Time",
    x = "Year",
    y = "Average Price (USD)",
    color = "Food Item"
  ) +
  theme_minimal()
```

This line plot shows how the average prices of selected food items changed over time. Overall, most food items show an upward trend, indicating that food prices generally increased over the years. However, the rate and pattern of increase differ across items. Ground beef shows the largest overall increase, while bananas remain relatively stable compared with the other items. Eggs show more noticeable fluctuations, suggesting that some food prices are more volatile than others.

```{r food-price-summary-table, echo=FALSE}
summary_table <- food_clean |>
  group_by(item_clean) |>
  summarise(
    earliest_date = min(date, na.rm = TRUE),
    latest_date = max(date, na.rm = TRUE),
    earliest_price = value[date == min(date, na.rm = TRUE)][1],
    latest_price = value[date == max(date, na.rm = TRUE)][1],
    absolute_change = latest_price - earliest_price,
    percent_change = ((latest_price - earliest_price) / earliest_price) * 100
  ) |>
  ungroup() |>
  mutate(
    earliest_price = round(earliest_price, 2),
    latest_price = round(latest_price, 2),
    absolute_change = round(absolute_change, 2),
    percent_change = round(percent_change, 1)
  )
```

```{r tbl-summary, echo=FALSE}
summary_table |>
  kable(
    caption = "Summary of Price Changes for Selected Food Items",
    col.names = c(
      "Food Item", "Earliest Date", "Latest Date",
      "Earliest Price", "Latest Price",
      "Absolute Change", "Percent Change (%)"
    ),
    align = "lcccccc"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center",
    latex_options = "scale_down"
  )
```

The summary table complements the line plot by presenting the earliest and latest observed prices for each food item, along with the absolute and percentage change over time. While the graph helps visualize long-term trends and short-term fluctuations, the table makes it easier to compare the overall magnitude of price changes across items.

```{r, include = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(kableExtra)
library(knitr)

eggs_data <- read.csv("Grade-A-Eggs-Price-NSA.csv", check.names = FALSE, stringsAsFactors = FALSE)

data_clean <- eggs_data |>
  mutate(Date = as.Date(paste0(Label, " 01"), format = "%Y %b %d")) |>
  mutate(Value = suppressWarnings(as.numeric(Value))) |>
  mutate(`1-Month % Change` = suppressWarnings(as.numeric(`1-Month % Change`)))
```

The line plot below demonstrates the Monthly Percent Change in the prices of Grade A Eggs from 2016 to 2026. The plot currently demonstrates only the monthly changes with 1 months sepearations, but this can be modified with 3-month, 6-month, and 12-month separations.

```{r, echo = FALSE, fig.width = 10, fig.height = 6, fig.cap = "Monthy Percent Change in Egg Prices", fig.alt = "Line Plot from 2016 to 2026"}
ggplot(data_clean |> filter(!is.na(`1-Month % Change`)),
                      aes(x = Date, y = `1-Month % Change`)) +
  geom_line(color = "red", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Grade A Eggs Price Over Time",
    subtitle = "1 Month Percent Change",
    x = "Year",
    y = "% Change"
  ) +
  theme(plot.title = element_text(face = "bold")) 
```

The following area chart shows the change in the actual prices of eggs over the past 10 years.

```{r, echo = FALSE, fig.width = 10, fig.height = 6, fig.cap = "Changes in Egg Prices", fig.alt = "Area Chart from 2016 to 2026"}
ggplot(data_clean |> filter(!is.na(Value)),
                          aes(x = Date, y = Value)) +
  geom_area(alpha = 0.75, fill = 'blue') +
  theme_minimal() +
  labs(
    title = "Grade A Eggs Price Over Time",
    x = "Year",
    y = "Price ($)"
  ) +
  theme_minimal()
```
As we can see from the plots above, the prices of eggs have generally increased as time goes on. There are a few spikes that stick out to take note of. One spike tends to occur in 2020 among all of the Percent Change plots. We can attribute these spikes to the COVID-19 pandemic and the shutdown that the world faced in early 2020. Some other notable spikes and drops that we can see are in early 2023 and early 2025. We can diagnose the reason by analyzing the seasonal spikes that occurred, which happen to be in winter. Winter is notorious for being flu season, which is exactly what caused  spikes [@schulz2023eggs]. In late 2022/24 to early 2023/25, the country faced an outbreak of avian flu, which would lead to the deaths of millions of egg-laying hens nationwide.

```{r, echo = FALSE}
summary_table <- data_clean |>
  summarise(
    Mean_Price = round(mean(Value, na.rm = TRUE), 2),
    Median_Price = round(median(Value, na.rm = TRUE), 2),
    Min_Price = round(min(Value, na.rm = TRUE), 2),
    Max_Price = round(max(Value, na.rm = TRUE), 2),
    Standard_Deviation = round(sd(Value, na.rm = TRUE), 2),
    Earliest_Price = round(first(Value), 2),
    Latest_Price = round(last(Value), 2),
    Total_Percent_Change = round((((Latest_Price - Earliest_Price)/Earliest_Price) * 100), 2)
  )
```

```{r, echo = FALSE}
summary_table |>
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "Metric",
    values_to = "Value"
  ) |>
  dplyr::mutate(
    Metric = dplyr::recode(Metric,
      Mean_Price = "Mean Price",
      Median_Price = "Median Price",
      Min_Price = "Minimum Price",
      Max_Price = "Maximum Price",
      Standard_Deviation = "Standard Deviation",
      Earliest_Price = "Earliest Price (Jan 2016)",
      Latest_Price = "Latest Price (Mar 2026)",
      Total_Percent_Change = "Percent Change"
    )
  ) |>
  knitr::kable(
    col.names = c("Metric", "Value"),
    align = "ll",
    caption = "Summary of Price Changes for Grade A Eggs"
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  )
```

The above summary table displays some interesting information about the Grade A Eggs dataset, including the earliest price from the time period selected (Jan 2016) and the most recent price (Mar 2026). With this, we can identify the total percent change in the price of eggs over the past 10 years to be 0.86%, meaning prices are only 0.86% higher in March 2026 than they were in January 2016.

\newpage

# Discussion / Conclusion
Overall, our analysis has revealed the relationship between food prices and overall inflation in the US on a regular basis. According to the data provided, the prices of most items have been increasing over time, which indicates a correlation between food prices and inflation, as expected. Although not all of the prices demonstrated the consistency of their changes, it is clear that the overall rise of their values shows how the prices for common products have been changing over time and affecting people.

However, one must keep in mind that even among similar products, their behaviour can differ from each other. For instance, ground beef demonstrated consistency when rising; banana prices have changed insignificantly; and eggs experienced drastic rises in price throughout several periods of time. Therefore, this allows concluding that inflation affects different goods differently, making some products especially vulnerable to changes in their production processes due to different factors.

As seen from the above analysis, inflation not only exists but is also evident in food prices. The latter have been changing in time, growing on average for each item. However, such a change does not affect all food items in the same manner, and it is possible to observe significant differences depending on the type of product under discussion.

# Author Contributions

Hanna led the cleaning and wrangling of the Average Price dataset, created key visualizations, and contributed to the written analysis.

Aria worked on the egg price dataset, including cleaning, visualization, and interpretation of short-term price fluctuations.

Aryan contributed to the development of visualizations, interpretation of findings, presentation preparation, and final editing of the report.

# Gen-AI Appendix

Quillbot was used in texts to check for Grammar and Spelling mistakes

Date: 6th May, 2026

Prompts: The paragraphs were given, and it was evaluated if any changes needed to be made

Response: Corrections such as punctuation and spelling

# Code Appendix

```{r code-Appendix, echo=TRUE, eval=FALSE}
#| label: code-Appendix
#| ref-label: !expr knitr::all_labels()
#| echo: true
#| eval: false

## Tidying the food prices data set

library(tidyverse)
library(knitr)
library(kableExtra)

food_clean <- read_csv("clean_food_prices.csv")

food_clean <- food_clean |>
  mutate(date = as.Date(date))

## Average Prices of Selected Food Items Over Time

ggplot(food_clean, aes(x = date, y = value, color = item_clean)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Average Prices of Selected Food Items Over Time",
    x = "Year",
    y = "Average Price (USD)",
    color = "Food Item"
  ) +
  theme_minimal()

## Summary Table of Price Changes for Selected Food items

summary_table <- food_clean |>
  group_by(item_clean) |>
  summarise(
    earliest_date = min(date, na.rm = TRUE),
    latest_date = max(date, na.rm = TRUE),
    earliest_price = value[date == min(date, na.rm = TRUE)][1],
    latest_price = value[date == max(date, na.rm = TRUE)][1],
    absolute_change = latest_price - earliest_price,
    percent_change = ((latest_price - earliest_price) / earliest_price) * 100
  ) |>
  ungroup() |>
  mutate(
    earliest_price = round(earliest_price, 2),
    latest_price = round(latest_price, 2),
    absolute_change = round(absolute_change, 2),
    percent_change = round(percent_change, 1)
  )

summary_table |>
  kable(
    caption = "Summary of Price Changes for Selected Food Items",
    col.names = c(
      "Food Item", "Earliest Date", "Latest Date",
      "Earliest Price", "Latest Price",
      "Absolute Change", "Percent Change (%)"
    ),
    align = "lcccccc"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center",
    latex_options = "scale_down"
  )

## Tidying Data for Monthly Percent Change in Egg Prices

library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(kableExtra)
library(knitr)

eggs_data <- read.csv("Grade-A-Eggs-Price-NSA.csv", check.names = FALSE, stringsAsFactors = FALSE)

data_clean <- eggs_data |>
  mutate(Date = as.Date(paste0(Label, " 01"), format = "%Y %b %d")) |>
  mutate(Value = suppressWarnings(as.numeric(Value))) |>
  mutate(`1-Month % Change` = suppressWarnings(as.numeric(`1-Month % Change`)))

## Line Plot for Monthly Percent Change in Egg Prices

ggplot(data_clean |> filter(!is.na(`1-Month % Change`)),
                      aes(x = Date, y = `1-Month % Change`)) +
  geom_line(color = "red", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Grade A Eggs Price Over Time",
    subtitle = "1 Month Percent Change",
    x = "Year",
    y = "% Change"
  ) +
  theme(plot.title = element_text(face = "bold"))

## Area Plot for Changes in Egg Prices

ggplot(data_clean |> filter(!is.na(Value)),
                          aes(x = Date, y = Value)) +
  geom_area(alpha = 0.75, fill = 'blue') +
  theme_minimal() +
  labs(
    title = "Grade A Eggs Price Over Time",
    x = "Year",
    y = "Price ($)"
  ) +
  theme_minimal()

## Summary of Price Changes for Grade A Eggs

summary_table <- data_clean |>
  summarise(
    Mean_Price = round(mean(Value, na.rm = TRUE), 2),
    Median_Price = round(median(Value, na.rm = TRUE), 2),
    Min_Price = round(min(Value, na.rm = TRUE), 2),
    Max_Price = round(max(Value, na.rm = TRUE), 2),
    Standard_Deviation = round(sd(Value, na.rm = TRUE), 2),
    Earliest_Price = round(first(Value), 2),
    Latest_Price = round(last(Value), 2),
    Total_Percent_Change = round((((Latest_Price - Earliest_Price)/Earliest_Price) * 100), 2)
  )

summary_table |>
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "Metric",
    values_to = "Value"
  ) |>
  dplyr::mutate(
    Metric = dplyr::recode(Metric,
      Mean_Price = "Mean Price",
      Median_Price = "Median Price",
      Min_Price = "Minimum Price",
      Max_Price = "Maximum Price",
      Standard_Deviation = "Standard Deviation",
      Earliest_Price = "Earliest Price (Jan 2016)",
      Latest_Price = "Latest Price (Mar 2026)",
      Total_Percent_Change = "Percent Change"
    )
  ) |>
  knitr::kable(
    col.names = c("Metric", "Value"),
    align = "ll",
    caption = "Summary of Price Changes for Grade A Eggs"
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  )
```
